Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-05-16 src/tex/ms.tex
[showyourwork] overleaf sync

    Using Maximum Likelihood Estimation (MLE) we could try to get an estimate $\hat{\mathbf{x}}$ directly, using some optimization procedure. However in general this represents an \textit{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}_1})$ look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation as well as the level of noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by minima. Depending on the initial estimate for $\mathbf{x}$ and the minimization algorithm the MLE based estimate is likely to only find such a local minimum. This is fundamentally the reason for RL decomposing estimates into point sources, described in the previous section.

    \todo{did you know this sum corresponds to the loop over data sets in machine learning...?}

@@

    In the case of non-overlapping patches the

    \begin{itemize}
        \item the case of non-overlapping patches is rather simple
        \item for overlapping patches we have to make a few more approximations
        \item either we neglect the correlation between overlapping patches
        \item or we have to optimize the array of $\mathbf{k}$ at the same time, instead of per patch optimization. However optimization on discrete 
        values is not easily possible in Pytorch. However here is one idea:
        group the GMM components by similarity and introduce a single continues
        parameter $z$, which interpolates between similar components.
        \item Another idea is to use Gumbel Softmax...
        \item we also compute the mean of the likelihood for the overlapping parts
    \end{itemize}

    To use the learned GMM as prior for the reconstruction of another image, the image is split into overlapping patches. For each of the patches the log-likelihood for each of the GMM components is evaluated. The resulting grid of overlapping patches is illustrated in Figure~\ref{fig:patches}.

    \begin{figure}
        \script{patches.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/patches.pdf}
            \caption{
                Grid of overlapping quadratic patches of size $8\times8$ pixels on top of an example image of the size $128 \times 68$ pixels. The overlap size was chosen to be $2$ pixels, which results in 50 patches in total. The patches are shown translucent, such that the light \enquote{stripes} result from two neighboring overlapping patches, while the small, even lighter squares result from four patches overlapping at the their corners.
            }
            \label{fig:patches}
        \end{centering}
    \end{figure}

    with radio astronomy data. They have shown that the reconstructed image only weakly depends on the choice of the reference data on which the the patch prior is learned. They found equivalent results for GMMs learned on natural images and specifically simulated images of black hole ring structures.

    Using half quadratic splitting and the Wiener filter solution, which is not possible in our case. Instead we choose a direct optimization by relying on differentiable programming later.

@@

    \subsection{Jolideco GMM Library}
    \label{ssec:jolideco-gmm-library}
    On radio data \cite{Bouman2016} found the results to be mostly independent of the choice of data the GMM prior was learned on. The strength of the GMM prior is its high \textit{expressiveness}, meaning the prior contains a large variety of patch shapes and adapts to the data in the optimization process. However in the low signal-to-noise regime we typically expect a higher dependence of the results on the choice of the data the GMM was learned on. The choice of the prior becomes more important and to avoid introducing a strong bias it should be adapted to the analysis scenario. For this reason and for purposes of testing we learned and provide a selection of GMMs (a \textit{Library}) to be used with the patch prior for download in a \textit{GitHub} repository\footnote{\url{https://github.com/jolideco/jolideco-gmm-library}}.

    \subsubsection{Zoran \& Weiss}
    As baseline reference we use the original GMM provided by \cite{Zoran2011}. The model was learned from the Berkeley image database \citep{Martin2001} which contains natural images and images of every day scenes and objects. The model was learned by splitting the training images into patches of size 8x8 pixels. This resulted in a training set of $2 \cdot 10^{6}$ patches, which were used to fit a GMM with 200 components. The model is provided by the authors as part of the source code for download\footnote{\url{https://people.csail.mit.edu/danielzoran/epllcode.zip}}. More details of how the model was learned are given in the original publication.

@@

    \subsubsection{JWST Cassiopeia A}
    The James Webb Space Telescope (JWST) is a large, infrared space telescope launched in late 2021. It delivers image data a very high spatial resolution as well as excellent signal to noise ratio. Both of these properties make JWST data an excellent choice for extracting patches for training a GMM. To limit the data reduction effort we relied on a promotional image of the SNR \textit{Cassiopeia A}\footnote{We downloaded the full resolution (4k)  PNG image from \url{https://webbtelescope.org/contents/media/images/2023/121/01GWQBBY77MHGFV3M3N63KDCEJ}}.
    We first converted the image to gray scale using a perceived brightness model \todo{reference skimage...}. Then we split the image into $approx 2$ million patches and learned a GMM with 128 mixture components.

@@

    \begin{itemize}
        \item Results are available in \url{https://github.com/jolideco/jolideco-benchmarks}
        \item We tried the binary search tree as proposed in \cite{Parameswaran2018},
        but without success, most likely because the native \texttt{Pytorch} support
        for parallelisation speeds up the standard GMM quite a bit...
        \item Could use a Neural Net for GMM component selection as proposed in \cite{Rosenbaum15}
        \item GPU support via \texttt{Pytorch}

@@

    \subsection{\fermi Event Types}
      \begin{figure*}
        \script{example-fermi-lat.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/example-fermi-lat.pdf}
            \caption{
                Illustration of the result of the \jolideco method applied to \fermi data of the supernova remnant \textit{RX J0852.0-4622} or \textit{Vela Junior}. The left image shows the counts above \qty[mode = text]{10}{GeV}. The different event classes are stacked into a single image. The image in the center shows the flux reconstructed by the \jolideco method. The image on the right shows the residuals as computed by Equation~\ref{eq:approx-sigma} and smoothed with a \textit{Gaussian} kernel of width 5~pixels. More information on this analysis example can be found on \url{https://github.com/jolideco/jolideco-fermi-examples}. \todo{Move color bar to top and make images larger}
            }
            \label{fig:example-fermi-lat}
        \end{centering}
    \end{figure*}
    As a second example we apply the \jolideco method to data from the Fermi Large Area Telescope (LAT). The LAT is a pair conversion \gammaray telescope on board of the Fermi satellite. It operates in the energy range between \qty[mode = text]{30}{MeV} to  \qty[mode = text]{300}{GeV}. For this example we chose again a Galactic source with complex morphology: the supernova remnant \textit{Vela Junior}. We use 14 years (from 2008 August 4 to 2023 August 4) of \texttt{P8R3\_SOURCE} photons with reconstructed energy in the range of \qty[mode = text]{10}{GeV} to  \qty[mode = text]{2}{TeV} and applied a zenith angle cut of \ang{105}. To preserve as much information as possible on the angular resolution of the instrument we split the data into the four event types \texttt{PSF0}, \texttt{PSF1}, \texttt{PSF2} and \texttt{PSF3}, as defined by the \fermi collaboration. We chose a pixel size of \qty[mode = text]{0.02}{$\deg$}, which corresponds to approximately 1/5th of the size of the PSF in the best event class. For the first part of the data reduction we use the standard \texttt{fermitools} \todo{add version and reference}.

>>>>

    Using Maximum Likelihood Estimation (MLE) we could try to get an estimate of $\hat{\mathbf{x}}$ directly, using some optimization procedure. However in general this represents an \textit{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}_1})$ look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation as well as the exitence of noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by minima. Depending on the initial estimate for $\mathbf{x}$ and the minimization algorithm the MLE based estimate is likely to only find such a local minimum. This is fundamentally the reason for RL decomposing estimates into point sources, described in the previous section.

@@

    \citep{Zoran2011} found that the reconstruction quality could be improved by working with a grid of overlapping patches. Such a grid is illustrated in Figure~\ref{fig:patches}. However using a grid of overlapping patches makes the interpretation of the prior as likelihood more difficult, because pixels at the corners and edges of patches are counted multiple times. For this we introduce a weighting mask when evaluating the likelihood of the GMM. This mask weights the pixels of the patch by their overlap with the neighboring patch. Overlapping edges are assigned a weight of 1/2, while overlapping corners are assigned a weight of 1/4. We keep the process of choosing the most likely component, however we note that in this case we neglected the correlation between neighbouring patches. As the choice of the GMM component per patch is discrete a simultaneous gradient based optimization of all patches is not easily possible, because we cannot compute gradients across GMM components.
    This would require a continuous representation or interpolation between GMM components.

    % \begin{itemize}
    %     \item the case of non-overlapping patches is rather simple
    %     \item for overlapping patches we have to make a few more approximations
    %     \item either we neglect the correlation between overlapping patches
    %     \item or we have to optimize the array of $\mathbf{k}$ at the same time, instead of per patch optimization. However optimization on discrete 
    %     values is not easily possible in Pytorch. However here is one idea:
    %     group the GMM components by similarity and introduce a single continues
    %     parameter $z$, which interpolates between similar components.
    %     \item Another idea is to use Gumbel Softmax...
    %     \item we also compute the mean of the likelihood for the overlapping parts
    % \end{itemize}

    % To use the learned GMM as prior for the reconstruction of another image, the image is split into overlapping patches. For each of the patches the log-likelihood for each of the GMM components is evaluated. The resulting grid of overlapping patches is illustrated in Figure~\ref{fig:patches}.

    \begin{figure}
        \script{patches.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/patches.pdf}
            \caption{
                Grid of overlapping quadratic patches of size $8\times8$~pixels on top of an example image of the size $128 \times 68$~pixels. The overlap size was chosen to be $2$ pixels, which results in 50 patches in total. The patches are shown translucent, such that the horizontal and vertical \enquote{stripes} result from two neighboring overlapping patches, while the small, even lighter squares result from four patches overlapping at the their corners.
            }
            \label{fig:patches}
        \end{centering}
    \end{figure}

    % with radio astronomy data. They have shown that the reconstructed image only weakly depends on the choice of the reference data on which the the patch prior is learned. They found equivalent results for GMMs learned on natural images and specifically simulated images of black hole ring structures.

    % Using half quadratic splitting and the Wiener filter solution, which is not possible in our case. Instead we choose a direct optimization by relying on differentiable programming later.

@@

    \subsection{Jolideco GMM Library}
    \label{ssec:jolideco-gmm-library}
    In general the strength of the GMM prior is its high \textit{expressiveness}. This means the prior contains a large variety of structural shapes and has a lot of freedom to adapt to the data in the optimization process. On radio data \cite{Bouman2016} found their results to be mostly independent of the choice of data the GMM prior was learned on. However in the low signal-to-noise regime we typically expect a higher dependence of the results on the choice of the data the GMM was learned on. The choice of the prior becomes more important and to avoid introducing a strong bias it should be adapted to the analysis scenario. For this reason and for purposes of testing we learned and provide a selection of GMMs (a \textit{Library}) to be used with the patch prior for download in a \textit{GitHub} repository\footnote{\url{https://github.com/jolideco/jolideco-gmm-library}}.

    \subsubsection{Zoran \& Weiss}
    As baseline reference we use the original GMM provided by \cite{Zoran2011}. The model was learned from the Berkeley image database \citep{Martin2001} which contains natural images and images of every day scenes and objects. The model was learned by splitting the training images into patches of size 8x8 pixels. This resulted in a training set of $2 \cdot 10^{6}$ patches, which were used to fit a GMM with 200 components. The model is provided by the authors as part of the source code for download\footnote{\url{https://people.csail.mit.edu/danielzoran/epllcode.zip}}. More details of how the model was learned are given in the original publication. This model uses the subtraction of the mean of the patch for normalisation.

@@

    \subsubsection{JWST Cassiopeia A}
    The James Webb Space Telescope (JWST) is a large, infrared space telescope launched in late 2021. It delivers image data at very high spatial resolution as well as excellent signal to noise ratio. Both of these properties make JWST data an excellent choice for extracting patches for training a GMM. To limit the data reduction effort we relied on a promotional image of the SNR \textit{Cassiopeia A}\footnote{We downloaded the full resolution (4k)  PNG image from \url{https://webbtelescope.org/contents/media/images/2023/121/01GWQBBY77MHGFV3M3N63KDCEJ}}.
    We first converted the image to gray scale using a perceived brightness model \todo{reference skimage...}. Then we split the image into $approx 2$ million patches and learned a GMM with 128 mixture components.

@@

    \begin{itemize}
        \item Results are available in \url{https://github.com/jolideco/jolideco-benchmarks}
        \item We tried the binary search tree as proposed in \cite{Parameswaran2018},
        but without success, most likely because the native \texttt{Pytorch} support
        for parallelisation speeds up the standard GMM quite a bit...
        \item Could use a Neural Net for GMM component selection as proposed in \cite{Rosenbaum15}
        \item GPU support via \texttt{Pytorch}
    \end{itemize}

@@

    \subsection{\fermi Event Types}
      \begin{figure*}
        \script{example-fermi-lat.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/example-fermi-lat.pdf}
            \caption{
                Illustration of the result of the \jolideco method applied to \fermi data of the supernova remnant \textit{RX J0852.0-4622} or \textit{Vela Junior}. The left image shows the counts above \qty[mode = text]{10}{GeV}. The different event classes are stacked into a single image. The image in the center shows the flux reconstructed by the \jolideco method. The image on the right shows the residuals as computed by Equation~\ref{eq:approx-sigma} and smoothed with a \textit{Gaussian} kernel of width 5~pixels. More information on this analysis example can be found on \url{https://github.com/jolideco/jolideco-fermi-examples}. \todo{Move color bar to top and make images larger}
            }
            \label{fig:example-fermi-lat}
        \end{centering}
    \end{figure*}
    As a second example we apply the \jolideco method to data from the Fermi Large Area Telescope (LAT). The LAT is a pair conversion \gammaray telescope on board of the Fermi satellite. It operates in the energy range between \qty[mode = text]{30}{MeV} to  \qty[mode = text]{300}{GeV}. For this example we chose again a Galactic source with complex morphology: the supernova remnant \textit{Vela Junior}. We use 14 years (from 2008 August 4 to 2023 August 4) of \texttt{P8R3\_SOURCE} photons with reconstructed energy in the range of \qty[mode = text]{10}{GeV} to  \qty[mode = text]{2}{TeV} and applied a zenith angle cut of \ang{105}. To preserve as much information as possible on the angular resolution of the instrument we split the data into the four event types \texttt{PSF0}, \texttt{PSF1}, \texttt{PSF2} and \texttt{PSF3}, as defined by the \fermi collaboration. We chose a pixel size of \qty[mode = text]{0.02}{$\deg$}, which corresponds to approximately 1/5th of the size of the PSF in the best event class. For the first part of the data reduction we use the standard \texttt{fermitools v2.2.0} \cite{}.
